---
title: PCA for tradeoff/synergy analysis and "variable importance"
author: Ben Schiek
date: '2020-07-18'
slug: pca-for-tradeoff-synergy-analysis-and-variable-importance
categories:
  - Methodology
tags: []
---

<style>
p.caption {
  font-size: 0.7em;
}
</style>

PCA typically used for...


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      warning = FALSE)

library(tidyverse)
setwd("C:/Users/bensc/OneDrive/Documents/Blog - Test")

this_folder <- "Data/"

#==============================================================================
# Define functions

# Barchart of variable-signal correlations
plot_corrXS_barchart <- function(mat_X_in, pct_explain = 90,
                                 n_signals_expln = NULL,
                                 n_signals_disp = 2,
                                 include_pctgExplnd = T,
                                 varimaxRot = T,
                                 group_info = NULL,
                                 sigNames = NULL){
  #----------------------------------------------------------------------
  # Set value for EITHER pct_explain or n_signals_expln, not both.
  # One or the other should be NULL.
  # (If both have values, pct_explain overrides.)
  # n_signals_disp is the number of signals to be displayed in graphic.
  #----------------------------------------------------------------------
  eig_values <- round(eigen(cor(mat_X_in))$values, 8)
  c_vec <- round(eig_values / sum(eig_values), 4)
  #----------------------------------------------------------------------
  if(!is.null(pct_explain)){
    ck_vec <- cumsum(c_vec)
    n_signals_expln <- which(ck_vec > 10^-2 * pct_explain)[1]
  }
  #----------------------------------------------------------------------
  mat_P <- eigen(cor(mat_X_in))$vectors
  if(mean(mat_P[, 1]) < 0){mat_P <- -mat_P}
  mat_G <- diag(eig_values)
  cormat_XS <- mat_P %*% sqrt(mat_G)
  row.names(cormat_XS) <- colnames(mat_X_in)
  mat_L <- cormat_XS
  #----------------------------------------------------------------------
  # Set sign of eigenvectors such that signals best conform to
  # their most highly correlated items
  #---
  # First have to get average of highest correlated items for each signal
  corrThresh <- 0.55
  n_items <- ncol(mat_L)
  list_X_hiCorr_avg <- list()
  for(i in 1:n_items){
    this_loadvec <- mat_L[, i]
    ind_tracks <- which(abs(this_loadvec) >= corrThresh)
    if(length(ind_tracks) == 0){
      ind_tracks <- which(abs(this_loadvec) == max(abs(this_loadvec)))
    }
    if(length(ind_tracks) == 1){
      list_X_hiCorr_avg[[i]] <- mat_X_in[, ind_tracks]
    }else{
      loadvec_kept <- this_loadvec[ind_tracks]
      list_X_hiCorr_avg[[i]] <- rowMeans(mat_X_in[, ind_tracks])
      
    }
  }
  mat_X_hiCorr_avg <- do.call(cbind, list_X_hiCorr_avg)
  mat_S_all <- mat_X_in %*% mat_P
  #---
  for(i in 1:n_items){
    this_S <- mat_S_all[, i]
    this_X_hiCorr_avg <- mat_X_hiCorr_avg[, i]
    mse <- mean((this_S - this_X_hiCorr_avg)^2)
    mse_neg <- mean((-this_S - this_X_hiCorr_avg)^2)
    if(mse_neg < mse){
      print("Changing eigenvector sign.")
      mat_P[, i] <- -mat_P[, i]
    }
  }
  #----------------------------------------------------------------------
  # cormat_XS <- mat_P %*% sqrt(mat_G)
  # row.names(cormat_XS) <- colnames(mat_X_in)
  # mat_L <- cormat_XS
  #----------------------------------------------------------------------
  # Get statistical significance info
  out_PCA <- FactoMineR::PCA(mat_X_in, scale.unit = T, ncp = n_signals_expln, graph = F)
  list_dimdesc <- FactoMineR::dimdesc(out_PCA,
                                      axes = c(1:n_signals_expln),
                                      proba = 1)
  list_matL <- list()
  list_matpVal <- list()
  for(i in 1:n_signals_expln){
    list_matL[[i]] <- plyr::join_all(list_dimdesc[[i]])[, 1]
    list_matpVal[[i]] <- plyr::join_all(list_dimdesc[[i]])[, 2]
  }
  mat_pVals <- do.call(cbind, list_matpVal)
  #---
  # Check loadings
  # mat_L_facto <- do.call(cbind, list_matL)
  # mat_L_facto[, 1] / mat_L[, 1]
  # out_PCA$var$coord[, 1] / mat_L[, 1]
  #-----------------------------------------------------------------------
  # Convert p-values to statistical significance scale between 0 and 1
  get_statSignifScale <- function(u){
    u <- log(1 / u)
    u <- (u - min(u)) / (max(u) - min(u))
    return(u)
  }
  #mat_statSignif <- apply(mat_pVals, 2, get_statSignifScale)
  mat_statSignif <- mat_pVals
  #-----------------------------------------------------------------------
  if(varimaxRot){
    mat_Lrot <- varimax(mat_L)[[1]]
    mat_Lrot <- matrix(as.numeric(mat_Lrot),
                       attributes(mat_Lrot)$dim,
                       dimnames = attributes(mat_Lrot)$dimnames)
    # mat_R <- varimax(mat_L)[[2]]
    # mat_R <- matrix(as.numeric(mat_R),
    #                 attributes(mat_R)$dim,
    #                 dimnames = attributes(mat_R)$dimnames)
    norm_m1to1 <- function(u){
      u <- 2 * (u - min(u)) / (max(u) - min(u)) - 1
      return(u)
      }
    for(i in 1:n_signals_disp){
      u <- mat_Lrot[, i]
      if(max(u) > 1.1 | min(u) < -1.1){
        mat_Lrot[, i] <- norm_m1to1(u)
      }
    }
    mat_L <- mat_Lrot
    xAxis_title <- "Correlations (Varimax Rotated)"
  }else{
    xAxis_title <- "Correlations"
  }
  #-----------------------------------------------------------------------
  df_L <- data.frame(Item = row.names(mat_L), mat_L[, 1:n_signals_disp])
  df_L$Item <- as.character(df_L$Item)
  df_statSignif <- data.frame(Item = row.names(mat_statSignif), mat_statSignif[, 1:n_signals_disp])
  df_statSignif$Item <- as.character(df_statSignif$Item)
  #-----------------------------------------------------------------------
  if(is.null(sigNames)){
    signal_id <- paste("PC", 1:n_signals_disp)
  }else{
    signal_id <- sigNames
  }
  #-------------------------------------------------------
  if(include_pctgExplnd){
    signal_id <- paste0(signal_id, "\n(", 10^2 * c_vec[1:n_signals_disp], "%)")
  }
  #-------------------------------------------------------
  colnames(df_L)[2:(n_signals_disp + 1)] <- signal_id
  colnames(df_statSignif)[2:(n_signals_disp + 1)] <- signal_id
  #-------------------------------------------------------
  gathercols <- as.character(signal_id)
  df_plot <- gather_(df_L, "Signal", "Correlation", gathercols)
  df_statSignif <- gather_(df_statSignif, "Signal", "Statistical\nsignificance", gathercols)
  df_plot <- merge(df_plot, df_statSignif, by = c("Item", "Signal"))
  df_plot$Signal <- factor(df_plot$Signal, levels = gathercols)
  if(!is.null(group_info)){
    outlist <- group_fn(group_info)
    cols_ordered_by_group <- outlist[[1]]
    group_color_vec <- outlist[[2]]
    group_vec_ordered <- outlist[[3]]
    df_match_group <- data.frame(Item = cols_ordered_by_group, Group = group_vec_ordered)
    df_plot <- merge(df_plot, df_match_group, by = "Item")
    df_plot <- df_plot[order(df_plot$Group), ]
    df_plot$Item <- factor(df_plot$Item, levels = unique(df_plot$Item))
    df_plot$`Statistical\nsignificance` <- round(df_plot$`Statistical\nsignificance`, 2)
    gg <- ggplot(df_plot, aes(x = Item, y = Correlation,
                              fill = Group))#,
                              #label = `Statistical\nsignificance`))#, alpha = `Statistical\nsignificance`))
    gg <- gg + scale_fill_manual(values = unique(group_color_vec))
  }else{
    gg <- ggplot(df_plot, aes(x = Item, y = Correlation, alpha = `Statistical\nsignificance`))
  }
  gg <- gg + geom_bar(stat = "identity", color = "grey", position = "dodge")
  #gg <- gg + geom_text(size = 4)
  gg <- gg + ylim(limits = c(-1, 1))
  gg <- gg + facet_wrap(~ Signal, nrow = 1)
  gg <- gg + labs(y = xAxis_title)
  gg <- gg + theme_bw()
  gg <- gg + theme(axis.text = element_text(size = 8),
                   axis.title.x = element_text(size = 8),
                   axis.title.y = element_blank(),
                   #legend.title = element_blank(),
                   legend.text = element_text(size = 8),
                   legend.title = element_blank(),
                   strip.text = element_text(size = 8))
  gg <- gg + coord_equal()
  gg <- gg + coord_flip()
  gg
  
}
#=====================================================================
#=====================================================================
# Define function to order data by group
group_fn <- function(group_info){
  list_groups <- group_info[[1]]
  group_names <- group_info[[2]]
  group_colors <- group_info[[3]]
  varNames_ordered <- do.call(c, list_groups)
  n_groups <- length(group_names)
  n_items <- length(varNames_ordered)
  if(is.na(group_colors)){
    bag_of_colors <- randomcoloR::distinctColorPalette(k = 5 * n_groups)
    group_colors <- sample(bag_of_colors, n_groups)
    #group_colors <- viridis::viridis_pal(option = "D")(length(group_names))
  }
  #if(reverse_order){group_colors <- rev(group_colors)}
  #varNames_ordered <- colnames(mat_pctDiff)
  group_vec <- rep(NA, n_items)
  group_color_vec <- rep(NA, n_items)
  for(i in 1:n_groups){
    this_group_vec <- list_groups[[i]]
    this_group_name <- group_names[i]
    this_group_color <- group_colors[i]
    group_vec[which(varNames_ordered %in% this_group_vec)] <- this_group_name
    group_color_vec[which(varNames_ordered %in% this_group_vec)] <- this_group_color
  }
  ind_ordered_cols <- order(factor(group_vec))
  cols_ordered_by_group <- as.character(varNames_ordered[ind_ordered_cols])
  group_color_vec <- group_color_vec[ind_ordered_cols]
  group_vec_ordered <- group_vec[ind_ordered_cols]
  out_list <- list(cols_ordered_by_group, group_color_vec, group_vec_ordered, ind_ordered_cols, group_vec)
  return(out_list)
}
#group_info <- list(list_groups, group_names, group_colors)
#==========================================================================

```

...Figure \@ref(fig:loadings)

```{r loadings, fig.show = "hold", fig.width = 7, fig.height=5, fig.align="center", fig.cap="Diets are diverse.", echo = FALSE}
# Data file created by AVISA_PCA.R
this_file <- "PCA FAO WDI etc data 2018.csv"
this_filepath <- paste0(this_folder, this_file)
df <- read.csv(this_filepath, stringsAsFactors = F, header = F)
df$V1 <- NULL
colnames(df) <- df[1, ]
df <- df[-1, ]
#----------------------------------------------------------------------------
#mat_X_in <- as.matrix(df[, -1])
mat_X_in <- apply(df[, -1], 2, as.numeric)
row.names(mat_X_in) <- df$Area
#----------------------------------------------------------------------------
stapleDemand_vec <- c("kcal/capita/day Animal Products", "kcal/capita/day Cereals", "kcal/capita/day Fruits/Veg.", "kcal/capita/day Oilcrops", "kcal/capita/day Pulses", "kcal/capita/day Starchy Roots", "kcal/capita/day Sugar & Sweeteners")
prodConsumPrice_vec <- c("Prod. Price Index (Avg. 2016-2018)",
                         "Prod. Price Volatility (CV)",
                         "Food CPI Volatility (CV) 2018")
landUse_vec <- c("Agricultural land (share)",
                 "Forest land (share)",
                 "Land under permanent crops (share)")
macro_vec <- c("Gross Domestic Product per capita",
               "Value Added (Agriculture, Forestry and Fishing)",
               "Value Added (Total Manufacturing)",
               "Consumption Inequality (Gini Coefficient)",
               "Income Inequality (Gini Coefficient)",
               "Export Unit/Value Index (2004-2006 = 100)",
               "Import Unit/Value Index (2004-2006 = 100)",
               "Total natural resources rents (% of GDP)",
               "Total natural resources rents (% of GDP)")
demographic_vec <- c("School enrollment, secondary (% gross)",
                     "Proportion of seats held by women in national parliaments (%)",
                     "Ratio of female to male labor force participation rate (%) (modeled ILO estimate)",
                     "Urban/Rural Pop.",
                     "Individuals using the Internet (% of population)",
                     "Life expectancy at birth, total (years)")
employment_vec <- c("Age dependency ratio, old (% of working-age population)",
                    "Age dependency ratio, young (% of working-age population)",
                    "Employment in agriculture (% of total employment) (modeled ILO estimate)",
                    "Employment in industry (% of total employment) (modeled ILO estimate)",
                    "Employment in services (% of total employment) (modeled ILO estimate)")

group_names <- c("Staple\nConsumption", "Producer and\nConsumer Prices",
                 "Land Use", "Macroeconomic\nIndicators",
                 "Demographics", "Employment")

list_groups <- list(stapleDemand_vec, prodConsumPrice_vec,
                    landUse_vec, macro_vec,
                    demographic_vec, employment_vec)
n_groups <- length(group_names)
bag_of_colors <- randomcoloR::distinctColorPalette(k = 2 * n_groups)
group_colors <- sample(bag_of_colors, n_groups)
group_info <- list(list_groups, group_names, group_colors)
#----------------------------------------------------------------------------

# eig_values <- round(eigen(cor(mat_X_in))$values, 8)
# #View(cor(mat_X_in))
# c_vec <- round(eig_values / sum(eig_values), 4)
# ck_vec <- cumsum(c_vec)
# df_plot <- data.frame(Signal = paste("Signal", 1:length(eig_values)), Portion_explained = c_vec, Portion_explained_cumul = ck_vec)
# colnames(df_plot)[2:3] <- c("Individually", "Cumulatively")
# gathercols <- colnames(df_plot)[2:3]
# df_plot <- gather_(df_plot, "Portion explained", "Value", gathercols)
# df_plot$Signal <- factor(df_plot$Signal,
#                          levels = unique(df_plot$Signal),
#                          ordered = T)
# gg <- ggplot(df_plot, aes(x = Signal, y = Value, fill = `Portion explained`))
# gg <- gg + geom_bar(stat = "identity", color = "black", position = "dodge")
# gg <- gg + scale_fill_manual(values = c("wheat3", "paleturquoise"))
# gg <- gg + geom_hline(yintercept = 0.9, color = "coral", size = 1, linetype = "dashed")
# gg <- gg + theme(axis.text.y = element_text(size = 7),
#                  axis.text.x = element_text(size = 7, angle = 60, hjust = 1),
#                  axis.title = element_blank(),
#                  legend.title = element_text(size = 8),
#                  legend.text = element_text(size = 8))
# gg
# #----------------------------------------------------------------------------
# n_signals <- which(ck_vec > 0.9)[1]
# n_signals
# #----------------------------------------------------------------------------
plot_corrXS_barchart(mat_X_in, pct_explain = 90,
                                 n_signals_expln = NULL,
                                 n_signals_disp = 3,
                                 include_pctgExplnd = T,
                                 varimaxRot = T,
                                 group_info = group_info,
                                 sigNames = NULL)
```

* Let's start with the second principal component (PC 2) since that's pretty easy. See how how only producer price variables are substantially correlated with it? Therefore you can say that PC 2 is mostly about producer prices. PC 1 has a lot more "loading onto" (i.e. correlated with) it--which seems to happen a lot with PC 1 in any dataset--but even so you can extract a meaning. If you establish 0.4 or 0.5 as a cut off point, that already weeds out a lot of variables as "not so important". The highest correlated variables with PC 1 are about economic growth--you've got GDP per capita plus human/social capital indicators (good education and healthcare, internet access), urbanization, and consumption of animal products and sweets, well known to go with rising incomes. So you could call PC 1 the "economic growth dimension". Note that employment in agriculture is negatively correlated with this economic growth dimension, while employment in the services sector is positively correlated, as one would expect. Note also that an inverted population pyramid goes with economic growth, while a not inverted population pyramid is negatively correlated with economic growth (as reflected in the "age dependency" variables). Again, no big surprise there. One thing that does strike me as interesting is that income inequality is negatively correlated with economic growth, as this contradicts the received wisdom that "a rising tide lifts all boats". And it is also interesting that natural resource rents are negatively correlated with economic growth. Together with the income inequality negative correlation, this can be taken to mean that capital intensive forms of growth (as opposed to labor intensive), which by their nature are socially exclusive, may ultimately be a drag on economic growth. This maybe raises a warning flag for places like Tanzania, whose growth is largely based on natural gas extraction, if I'm not mistaken.

* Instead of trying to interpret a gargantuan correlation matrix

* This is linear. For second order relations might extract variable importance from ML algorithms like random forest and xgboost.

* (In a talk at CIAT about a year ago, then Director Reuben Echeverria said...) The tradeoff aspect -- development agencies often talk about goals, but neglect to address or even acknowledge that tradeoffs may exist between goals. Here is a way to do that.