---
title: A closer look at "Bennett's Law"
author: Ben Schiek
date: '2020-07-12'
slug: a-closer-look-at-bennett-s-law
categories:
  - Nutrition transition
tags: []
bibliography: Bennetts Law blog post.bib
---

<style>
p.caption {
  font-size: 0.7em;
}
</style>


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      warning = FALSE)
library(plyr)
library(tidyverse)
library(patchwork)
library(ggrepel)
library(gganimate)
#library(kableExtra)

this_folder <- "Data/Bennetts Law data/"

shape_vec <- c(21:24, 4)
point_size <- 2
label_size <- 4

#==========================================================================
shift_legend2 <- function(p) {
  # ...
  # to grob
  gp <- ggplotGrob(p)
  facet.panels <- grep("^panel", gp[["layout"]][["name"]])
  empty.facet.panels <- sapply(facet.panels, function(i) "zeroGrob" %in% class(gp[["grobs"]][[i]]))
  empty.facet.panels <- facet.panels[empty.facet.panels]
  
  # establish name of empty panels
  empty.facet.panels <- gp[["layout"]][empty.facet.panels, ]
  names <- empty.facet.panels$name
  # example of names:
  #[1] "panel-3-2" "panel-3-3"
  
  # now we just need a simple call to reposition the legend
  #lemon::reposition_legend(p, 'center', panel=names)
  p_out <- lemon::reposition_legend(p, 'center', panel=names)
  #class(lemon::reposition_legend(p, 'center', panel=names))
  return(p_out)
}
#==========================================================================
```

"The broad fact that variety rather than uniformity is characteristic of food consumption throughout the world is familiar and indisputable. Differences are found in the 'dishes' consumed from country to country, region to region, locality to locality, even family to family. Within a country or region, these differences appear also from one epoch to another. They are so prevalent and conspicuous, in fact, that it seems almost hopeless to discover, in the international sphere, some way of characterizing summarily the diets of different nations and contrasting them one with the other" [@bennett1941international].

Even within a fairly small geographical region of Africa, diets can vary considerably (Figure \@ref(fig:dietDiversity)). If we were to look at each of these food catgories in more detail, we would see even more diversity in terms of what specific crops constitute the groups.

```{r dietDiversity, fig.show = "hold", fig.width = 7, fig.height=4, fig.align="center", fig.cap="Diets are diverse.", echo = FALSE}
this_file <- "Bennetts Law post_diverse diets.csv"
this_filepath <- paste0(this_folder, this_file)
df_plot <- read.csv(this_filepath, stringsAsFactors = F)
df_plot$X.1 <- NULL
df_plot$X <- NULL
df_plot <- df_plot[, c("Area", "Year", "Item", "kcal.share", "mu_food_area_item")]
colnames(df_plot)[4] <- "Share of daily diet"
df_plot$Item[grep("Sugar", df_plot$Item)] <- "Sugar\n& sweeteners"
df_plot$Item[grep("Animal", df_plot$Item)] <- "Animal\nproducts"
df_plot$Item[grep("Roots", df_plot$Item)] <- "Starchy\nroots"
#-----------------------------------------------------------------------------
area_vec <- c("Kenya", "Uganda", "Ethiopia", "Tanzania", "Rwanda")
df_plot <- subset(df_plot, Area %in% area_vec &
                    Year > 1992)
df_plot$Year <- as.integer(df_plot$Year)
yr_min <- min(df_plot$Year)
yr_max <- max(df_plot$Year)
n <- length(unique(df_plot$Item))
bag_of_colors <- randomcoloR::distinctColorPalette(k = 2 * n)
color_vec <- sample(bag_of_colors, n)
#good_colors <- c("#B346E2", "#74B4D0", "#E16953", "#D6E1D8", "#C7E84D", "#6BE47C", "#E4B64D")
#color_vec <- good_colors
df_plot <- df_plot %>% group_by(Area, Item) %>%
  mutate(mu_food_area_item = mean(`Share of daily diet`, na.rm = T)) %>%
  as.data.frame()
df_plot$Item <- factor(df_plot$Item,
                       levels = unique(df_plot$Item[order(df_plot$mu_food_area_item, df_plot$Item, decreasing = T)]),
                       ordered = T)
gg <- ggplot(df_plot, aes(x = Year, y = `Share of daily diet`, fill = Item))
gg <- gg + geom_area(position = "stack")
gg <- gg + scale_x_continuous(breaks = seq(yr_min, yr_max, length.out = 4))
gg <- gg + scale_fill_manual(values = color_vec)
gg <- gg + facet_wrap(~Area, ncol = 3)
gg <- gg + labs(caption = "Source: FAO")
gg <- gg + theme(legend.title = element_blank(),
                 axis.title.x = element_blank())
gg <- gg + guides(fill = guide_legend(nrow = 4, byrow = T))
gg <- shift_legend2(gg)
gg <- ggplotify::as.ggplot(gg)


```


## Bennett's Law

4 calories in a gram of protein or carbohydrate, 9 calories in a gram of fat.
https://www.nal.usda.gov/fnic/how-many-calories-are-one-gram-fat-carbohydrate-or-protein

...Figure \@ref(fig:Bennett0). This means that the protein share of the diet remains fairly constant, although protein sources may vary--passing from legumes to animal proteins, for example. The correlation with higher income can also be viewed in this variation by the size of the points.

```{r Bennett0, fig.show = "hold", fig.width = 6, fig.height=5, fig.align="center", fig.cap="Bennett's Law, FAO data for 2017.", echo = FALSE}
this_file <- "Bennetts Law data.csv"
this_filepath <- paste0(this_folder, this_file)
df_bennett <- read.csv(this_filepath, stringsAsFactors = F)
#colnames(df_bennett)
df_bennett$X <- NULL
#colnames(df_bennett) <- gsub(".", "\\", colnames(df_bennett), fixed = T)
colnames(df_bennett)[3:9] <- c("Food supply (kcal/capita/day)",
                               "Fat supply quantity (kcal/capita/day)",
                               "Protein supply quantity (kcal/capita/day)",
                               "Carb supply quantity (kcal/capita/day)",
                               "Fat share of diet (%)",
                               "Carb share of diet (%)",
                               "Carb share of diet (%), logged")
colnames(df_bennett)[11] <- "GDP / capita (USD), logged"
#-----------------------------------------------------------------------------
df_plot <- subset(df_bennett, Year == "2017")
df_mod <- df_plot[, c("Carb share of diet (%), logged", "GDP / capita (USD), logged")]
mod <- lm(`Carb share of diet (%), logged` ~., data = df_mod)
#summary(mod)
m <- round(mod$coefficients[2], 2)
b <- round(mod$coefficients[1], 2)
#-----------------------------------------------------------------------------
label_Africa <- c("Tanzania", "Uganda", "Ethiopia", "South Africa",
                  "Ghana")
label_NAfricaWAsia <- c("Turkey", "Saudi Arabia")
label_Asia <- c("India", "Iran", "China, mainland", "Vietnam", "Japan",
                "South Korea")
label_NAmEurAusNZ <- c("Italy", "France", "Australia")
label_LAC <- c("Nicaragua", "Colombia", "Brazil")
labelThese_vec <- c(label_Africa, label_NAfricaWAsia, label_Asia,
                    label_NAmEurAusNZ, label_LAC)
#-----------------------------------------------------------------------------

df_plot$label_these <- NA
u <- df_plot$Country
df_plot$label_these[which(u %in% labelThese_vec)] <- df_plot$Country[which(u %in% labelThese_vec)]

n <- length(unique(df_plot$Region))
bag_of_colors <- randomcoloR::distinctColorPalette(k = 2 * n)
color_vec <- sample(bag_of_colors, n)

gg <- ggplot(df_plot, aes(x = `GDP / capita (USD), logged`,
                          y = `Carb share of diet (%), logged`,             
                          group = Region, fill = Region,
                          shape = Region,
                          label = label_these))
# gg <- gg + geom_point(aes(fill = Region, group = Region, shape = Region), pch = 21, alpha = 0.5, size = 6, color = "black")
gg <- gg + geom_smooth(aes(group = NULL, fill = NULL, shape = NULL), method = lm, se = F)
gg <- gg + geom_point(alpha = 0.6, size = point_size, color = "black", stroke = 1)
gg <- gg + scale_fill_manual(values = color_vec)
gg <- gg + scale_shape_manual(values = shape_vec)
# gg <- gg + geom_abline(slope = m, intercept = b,
#                        linetype = "dashed", size = 1)
gg <- gg + geom_text_repel(color = "black", size = label_size)
gg <- gg + labs(#title = "Engel's Law", 
  subtitle = paste0("Slope = ", m, ", y-intercept = ", b))#,
#caption = "Source: Author's creation based on World Bank International Comparison Program Data for 2017.
#GDP/capita is based on 2015 prices.")
gg <- gg + guides(fill = guide_legend(nrow = 2, byrow = T, override.aes = list(linetype = 0)),
                  color = guide_legend(override.aes = list(linetype = 0)))
gg <- gg + theme(legend.position = "bottom",
                  legend.spacing.x = unit(0.25, 'cm'),
                  legend.title = element_blank())
gg

```

...Figure \@ref(fig:Bennett1). The shapes are sized accoridng to GDP/capita so that the correlation with income can be observed too.

```{r Bennett1, fig.show = "hold", fig.width = 6, fig.height=5, fig.align="center", fig.cap="A variation of Bennett's Law, FAO data for 2017.", echo = FALSE}

#------------------------------------------------------------------------
df_mod <- df_plot[, c("Carb share of diet (%)", "Fat share of diet (%)")]
mod <- lm(`Carb share of diet (%)` ~., data = df_mod)
#summary(mod)
m <- round(mod$coefficients[2], 2)
b <- round(mod$coefficients[1], 2)

# df_plot$Country[which(df_plot$`Fat share of diet (%)` ==
#                         max(df_plot$`Fat share of diet (%)`))]
df_plot <- subset(df_plot, Country != "Slovakia")
#colnames(df_plot)
df_plot$`GDP / capita (USD)` <- exp(df_plot$`GDP / capita (USD), logged`)
gg <- ggplot(df_plot, aes(x = `Fat share of diet (%)`,
                          y = `Carb share of diet (%)`,
                          group = Region, fill = Region,
                          shape = Region, size = `GDP / capita (USD)`,
                          label = label_these))
gg <- gg + geom_smooth(aes(group = NULL, fill = NULL, shape = NULL), method = lm, se = F)
gg <- gg + geom_point(alpha = 0.6, color = "black", stroke = 1)
gg <- gg + scale_fill_manual(values = color_vec)
gg <- gg + scale_shape_manual(values = shape_vec)
# gg <- gg + geom_abline(slope = m, intercept = b,
#                        linetype = "dashed", size = 1)
gg <- gg + geom_text_repel(color = "black", size = label_size)
gg <- gg + labs(#title = "Bennett's Law", 
  subtitle = paste0("Slope = ", m, ", y-intercept = ", b))#,
#caption = "Source: Author's creation based on World Bank International Comparison Program Data for 2017.
#GDP/capita is based on 2015 prices.")
gg <- gg + theme(legend.position = "bottom",
                legend.spacing.x = unit(0.25, 'cm'),
                legend.title = element_blank())
gg <- gg + guides(fill = guide_legend(nrow = 2, byrow = T, override.aes = list(linetype = 0)),
                  color = guide_legend(override.aes = list(linetype = 0)),
                  size = F)
gg
# gg1 + gg2 + plot_layout(ncol = 2, guides = "collect") & theme(legend.position = 'bottom')

```

Animation 1970-2018...
```{r EngelAnim, dev="png"}

df_plot <- df_bennett

df_plot$label_these <- NA
u <- df_plot$Country
df_plot$label_these[which(u %in% labelThese_vec)] <- df_plot$Country[which(u %in% labelThese_vec)]

df_plot <- subset(df_plot, Country != "Slovakia")
#colnames(df_plot)

df_plot$`GDP / capita (USD)` <- exp(df_plot$`GDP / capita (USD), logged`)
gg <- ggplot(df_plot, aes(x = `Fat share of diet (%)`,
                          y = `Carb share of diet (%)`,
                          group = Region, fill = Region,
                          shape = Region, size = `GDP / capita (USD)`,
                          label = label_these))
gg <- gg + geom_smooth(aes(group = NULL, fill = NULL, shape = NULL), method = lm, se = F)
gg <- gg + geom_point(alpha = 0.6, color = "black", stroke = 1)
gg <- gg + scale_fill_manual(values = color_vec)
gg <- gg + scale_shape_manual(values = shape_vec)
gg <- gg + geom_text(color = "black", size = label_size)
gg <- gg + labs(title = "Year: {frame_time}",
                subtitle = "Bennett's Law (variation thereof)")
#caption = "Source: Author's creation based on World Bank International Comparison Program Data for 2017.
#GDP/capita is based on 2015 prices.")
gg <- gg + theme(legend.position = "bottom",
                legend.spacing.x = unit(0.25, 'cm'),
                legend.title = element_blank())
gg <- gg + guides(fill = guide_legend(nrow = 2, byrow = T, override.aes = list(linetype = 0)),
                  color = guide_legend(override.aes = list(linetype = 0)),
                  size = F)
gg <- gg + transition_time(Year) + ease_aes("linear")
#gg
animate(gg, fps = 5)


```

...Figure \@ref(fig:Bennett0Stability)

```{r Bennett0Stability, fig.show = "hold", fig.width = 5, fig.height=4, fig.align="center", fig.cap="Bennett's Law (variation thereof) stability over time.", echo = FALSE}

df_params <- df_bennett[, c("Year", "Carb share of diet (%), logged",
                            "GDP / capita (USD), logged")]
df_params <- df_params[which(!is.na(df_params$`Carb share of diet (%), logged`)), ]

yr_vec <- unique(df_params$Year)
n_yrs <- length(yr_vec)
m_vec <- c()
b_vec <- c()
adjR2_vec <- c()
for(i in 1:n_yrs){
  this_year <- yr_vec[i]
  this_df <- subset(df_params, Year == this_year)
  this_df$Year <- NULL
  mod <- lm(`Carb share of diet (%), logged`~., this_df)
  m_vec[i] <- round(mod$coefficients[2], 3)
  b_vec[i] <- round(mod$coefficients[1], 3)
  df_out <- as.data.frame(broom::glance(mod))
  adjR2_vec[i] <- df_out$adj.r.squared
  
}

df_plot <- data.frame(Year = yr_vec, Slope = m_vec, b_vec, adjR2_vec)
colnames(df_plot)[3:4] <- c("Y-intercept", "Adj. R squared")
# mean(df_plot$Slope)
# sd(df_plot$Slope)
# mean(df_plot$`Y-intercept`)
# sd(df_plot$`Y-intercept`)

df_plot$Year <- as.integer(as.character(df_plot$Year))
df_plot <- df_plot %>% gather(Type, Value, Slope:`Adj. R squared`)
gg <- ggplot(df_plot, aes(x = Year, y = Value))
gg <- gg + geom_line()
gg <- gg + scale_x_continuous(breaks = seq(yr_vec[1], yr_vec[n_yrs - 1], length.out = 24))
gg <- gg + facet_wrap(~Type, ncol = 1, scales = "free_y")
gg <- gg + theme(axis.title = element_blank(),
                  axis.text.x = element_text(angle = 60, hjust = 1),
                  strip.background = element_blank())
gg


```

References