---
title: A "new" empirical law?
author: Ben Schiek
date: '2020-07-14'
slug: a-new-empirical-law
categories:
  - Empirical laws
tags: []
---

<style>
p.caption {
  font-size: 0.7em;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      warning = FALSE)
#devtools::install_github("thomasp85/patchwork")
library(plyr)
library(tidyverse)
library(zoo)
library(tidyquant)
library(patchwork)
library(ggrepel)
#library(kableExtra)
setwd("C:/Users/bensc/OneDrive/Documents/Blog - Test/content/post/")
this_folder <- "Data/New empirical law data/"
#===========================================================================
# Define function
shift_legend2 <- function(p) {
  # ...
  # to grob
  gp <- ggplotGrob(p)
  facet.panels <- grep("^panel", gp[["layout"]][["name"]])
  empty.facet.panels <- sapply(facet.panels, function(i) "zeroGrob" %in% class(gp[["grobs"]][[i]]))
  empty.facet.panels <- facet.panels[empty.facet.panels]
  
  # establish name of empty panels
  empty.facet.panels <- gp[["layout"]][empty.facet.panels, ]
  names <- empty.facet.panels$name
  # example of names:
  #[1] "panel-3-2" "panel-3-3"
  
  # now we just need a simple call to reposition the legend
  #lemon::reposition_legend(p, 'center', panel=names)
  p_out <- lemon::reposition_legend(p, 'center', panel=names)
  #class(lemon::reposition_legend(p, 'center', panel=names))
  return(p_out)
}
#===========================================================================

```

...Figure \@ref(fig:priceSeries). It seems that the higher the price, the more volatile it is... The phenomenon can be observed across a wide range of food and non-food agricultural commodities. It can also be seen to occur with both raw and processed versions of the commodities. The relation appears to persist across a wide range of food price epochs, from the oil driven spike of the 1970s through to the more recent 2007-2008 crisis. Plotting the mean price against its standard deviation offers striking confirmation... \@ref(fig:riskReward)

```{r priceSeries, fig.show = "hold", fig.width = 6, fig.height=4, fig.align="center", fig.cap="The higher the price, the more volatile. FAO data 1972-2017.", echo = FALSE}
this_file <- "FAO export data world.csv"
this_filepath <- paste0(this_folder, this_file)
df_export_world <- read.csv(this_filepath, stringsAsFactors = F)
df_export_world$X <- NULL
colnames(df_export_world)[4:6] <- c("Export Quantity", "Export Value",
                                    "World Export Price (USD / metric ton)")
#=============================================================================
plot_these <- c("Maize", "Apples", "Cocoa, beans", "Cocoa, butter")
df_plot <- subset(df_export_world, Item %in% plot_these &
                    Year >= 1972)
yr_1 <- unique(df_plot$Year)[1]
yr_n <- max(unique(df_plot$Year))
gg <- ggplot(df_plot, aes(x = Year, y = `World Export Price (USD / metric ton)`, group = Item, color = Item)) + geom_line(lwd = 1)
gg <- gg + scale_x_continuous(breaks = seq(yr_1, yr_n, length.out = 6))
#gg <- gg + scale_y_log10()
gg <- gg + theme(axis.title.x = element_blank(),
                 legend.title = element_blank(),
                 legend.position = "top")
gg

```

...Figure \@ref(fig:riskReward)

```{r riskReward, fig.show = "hold", fig.width = 6, fig.height=6, fig.align="center", fig.cap="Logged expected price plotted against logged standard deviation. 'Export price' here equals FAO's Export Value series divided by its Export Quantity series, downloaded from FAOSTAT.", echo = FALSE}

df_riskReward <- df_export_world %>% group_by(Item) %>%
  summarise(`Mean World Export Price (USD / metric ton)` = mean(`World Export Price (USD / metric ton)`, na.rm = T),
            `Stand. Dev. World Export Price (USD / metric ton)` = sd(`World Export Price (USD / metric ton)`, na.rm = T),
            `Mean Export Qty.` = mean(`Export Quantity`, na.rm = T), 
            `Mean Export Value ('000 USD)` = mean(`Export Value`, na.rm = T))

df_riskReward$`Mean World Export Price (USD / metric ton), logged` <-
  log(df_riskReward$`Mean World Export Price (USD / metric ton)`)
df_riskReward$`Stand. Dev. World Export Price (USD / metric ton), logged` <-
  log(df_riskReward$`Stand. Dev. World Export Price (USD / metric ton)`)
#--
df_riskReward <- df_riskReward[which(!is.na(df_riskReward$`Stand. Dev. World Export Price (USD / metric ton), logged`)), ]
#nrow(df_riskReward)
#----------------------------------------------------------------------------
df_mod <- df_riskReward[, c("Mean World Export Price (USD / metric ton), logged", "Stand. Dev. World Export Price (USD / metric ton), logged")]
#nrow(df_mod)
row.names(df_mod) <- df_riskReward$Item
mod <- lm(`Mean World Export Price (USD / metric ton), logged` ~., df_mod)
#summary(mod)
b <- round(mod$coefficients[1], 2)
m <- round(mod$coefficients[2], 2)
#---
# gg <- ggplot(data.frame(Item = row.names(df_mod), yhat = mod$fitted.values,
#                         resid = mod$residuals),
#              aes(x = yhat, y = resid, label = Item))
# gg <- gg + geom_point() + geom_text()
# gg <- gg + geom_hline(yintercept = c(-1, 1), color = "red")
# gg
#---
ind_outlier <- which(abs(as.numeric(mod$residuals)) >= 0.75)
excluded_items <- row.names(df_mod)[ind_outlier]
pct_removed <- 100 * length(ind_outlier) / nrow(df_riskReward)
row_names <- row.names(df_mod)[-ind_outlier]
df_mod <- df_mod[-ind_outlier, ]
row.names(df_mod) <- row_names
mod <- lm(`Mean World Export Price (USD / metric ton), logged` ~., df_mod)
#summary(mod)
b <- round(mod$coefficients[1], 2)
m <- round(mod$coefficients[2], 2)
#plot(mod$fitted.values, mod$residuals)
#----------------------------------------------------------------------------
#df_riskReward$Item[ind_outlier]
df_plot <- df_riskReward[-ind_outlier, c("Item", "Mean World Export Price (USD / metric ton), logged", "Stand. Dev. World Export Price (USD / metric ton), logged",
                             "Mean Export Value ('000 USD)")]

df_plot$`Mean Export Value\n(million USD)` <- df_plot$`Mean Export Value ('000 USD)` / 1000

labelItems <- c("Maize", "Cassava dried", "Flour, wheat", "Starch, cassava",
              "Coffee, green", "Coffee, roasted", "Wheat", "Bananas",
              "Plantains", "Potatoes", "Cassava Equivalent", "Cocoa, beans",
              "Wool, greasy", "Beans, dry", "Cotton lint", "Silk", "Apples",
              "Soybeans", "Oil, soybean", "Rubber, natural", "Rice",
              "Cocoa, butter")

df_plot$label_these <- NA
u <- df_plot$Item
df_plot$label_these[which(u %in% labelItems)] <- u[which(u %in% labelItems)]

label_size <- 4
gg <- ggplot(df_plot, aes(x = `Stand. Dev. World Export Price (USD / metric ton), logged`,
                          y = `Mean World Export Price (USD / metric ton), logged`))
gg <- gg + geom_smooth(method = lm, se = F)
gg <- gg + geom_point(aes(size = `Mean Export Value\n(million USD)`),
                      fill = "khaki1", pch=21, color="Black")
gg <- gg + geom_text_repel(aes(label = label_these), size = label_size)
gg <- gg + labs(title = "The risk-return frontier", subtitle = paste0("Slope = ", m, "  Y intercept = ", b))
gg <- gg + theme(legend.position = "bottom")
gg <- gg + labs(title = "The ag commodity risk-return frontier",
                caption = "Source: Author's creation using FAOSTAT export value and quantity series 1961-2017.")
gg
#ggsave("risk reward frontier.pdf")
```

...Figure \@ref(fig:riskRewardByRegion)

```{r riskRewardByRegion, fig.show = "hold", fig.width = 7, fig.height=6, fig.align="center", fig.cap="Logged expected price plotted against logged standard deviation, by region.", echo = FALSE}
this_file <- "FAO export data regions.csv"
this_filepath <- paste0(this_folder, this_file)
df_export_regions <- read.csv(this_filepath, stringsAsFactors = F)
df_export_regions$X <- NULL
colnames(df_export_regions)[4:6] <- c("Export Quantity", "Export Value",
                                    "Region Export Price (USD / metric ton)")
#============================================================================
#colnames(df_export_regions)
df_riskReward_regions <- df_export_regions %>% group_by(Region, Item) %>%
  summarise(`Mean Region Export Price (USD / metric ton)` = mean(`Region Export Price (USD / metric ton)`, na.rm = T),
            `Stand. Dev. Region Export Price (USD / metric ton)` = sd(`Region Export Price (USD / metric ton)`, na.rm = T),
            `Mean Export Qty.` = mean(`Export Quantity`, na.rm = T), 
            `Mean Export Value ('000 USD)` = mean(`Export Value`, na.rm = T))

df_riskReward_regions$`Mean Region Export Price (USD / metric ton), logged` <-
  log(df_riskReward_regions$`Mean Region Export Price (USD / metric ton)`)
df_riskReward_regions$`Stand. Dev. Region Export Price (USD / metric ton), logged` <-
  log(df_riskReward_regions$`Stand. Dev. Region Export Price (USD / metric ton)`)
df_riskReward_regions$`Mean Export Value\n(million USD)` <-
  df_riskReward_regions$`Mean Export Value ('000 USD)` / 1000
#--
df_riskReward_regions <- df_riskReward_regions[which(!is.na(df_riskReward_regions$`Stand. Dev. Region Export Price (USD / metric ton), logged`)), ]
ind_rm <- which(is.infinite(df_riskReward_regions$`Stand. Dev. Region Export Price (USD / metric ton), logged`))
#df_riskReward_regions$Item[ind_rm] #look
df_riskReward_regions <- df_riskReward_regions[-ind_rm, ]
#nrow(df_riskReward_regions)
#----------------------------------------------------------------------------
df_plot <- df_riskReward_regions
gg <- ggplot(df_plot, aes(x = `Stand. Dev. Region Export Price (USD / metric ton), logged`,
                          y = `Mean Region Export Price (USD / metric ton), logged`))
#gg <- gg + geom_smooth(method = lm, se = F)
gg <- gg + geom_point(aes(size = `Mean Export Value\n(million USD)`),
                      fill = "khaki1", pch=21, color="Black")
#gg <- gg + geom_text_repel(aes(label = label_these), size = label_size)
gg <- gg + facet_wrap(~Region)
gg <- gg + labs(title = "The ag commodity risk-return frontier, by region",
                caption = "Source: Author's creation using FAOSTAT export value and quantity series 1961-2017.")
gg <- shift_legend2(gg)
gg <- ggplotify::as.ggplot(gg)
```